Wrangling multivariate spatio-temporal data with the R package cubble

### 2022-04-30 3-advanced.Rmd
Advanced section-morework needed

Spatial locations can have grouping structures either inherent in the data or obtained by clustering. Rather than analysing variables in the site level, summarised variables in the cluster level can give a crisper picture of local areas. In cubble, \code{switch_key()} can be used to create a new level of grouping of spatial locations by specifying a clustering variable. Figure \ref{fig:illu-hier} illustrates the relationship of cubbles at station and cluster level, in both the long and nested form. By specifying \code{cluster_nested <- station_nested \%>\% switch_key(key = cluster)}, the cubble redefines the cubble key from the \code{id} in \code{station_nested} to \code{cluster} in \code{cluster_nested}. All the spatial variables variant to \code{cluster} are now nested into a \code{.val} column and cluster level variables can be computed in the same fashion as station level variables in \code{station_nested}.

@@

One task that may interest analysts in spatio-temporal data is to find how similar the time series from nearby sites are. This problem can be seen as a matching problem [@stuart2010matching; @mcintosh2018using] that pairs up similar time series in nearby locations or a data fusing exercise that merges data collected from different sources [@cocchi2019data]. \code{match_sites()} in \pkg{cubble} provides a simple algorithm for this task. The algorithm first matches the two data sources spatially by computing the pairwise distance on latitude and longitude. Pairs that pass the spatial matching are then matched temporally through computing the number of matched peaks within a fixed length window. Figure \ref{fig:illu-matching} illustrates this temporal matching in more details. Given two series \code{A} and \code{a}, three peaks have been picked in each series. An interval, with default length of five, is constructed for each peak in series \code{A} and the peaks in series \code{a} are tested against whether they fall into any of the intervals. In this illustration, there are two matches for these two series. Several arguments are available in \code{match_sites()} to fine-tune the matching:

@@

The linking structure of two forms in cubble fits naturally with the interactive graphic pipeline discussed in the literature [@buja1988elements; @buja1996interactive;  @sutherland2000orca; @xie2014reactive; @cheng2016enabling]. Diagram \ref{fig:illu-interactive} illustrates how linking works from the map to the time series in cubble. The map and time series plot is associated with the nested or long cubble, respectively, and when a user action is captured on the map, the site will be activated in the nested cubble (left). The nested cubble will communicate to the long cubble to activate all the observations with the same \code{id} (middle). The long cubble will then highlight the activated series in the time series plot (right).

>>>>

Spatial locations can have grouping structures either inherent to the data or obtained by clustering. In those cases, rather than analysing variables at the site level, it might be of interest to summarised variables at a cluster or group level, as that can give a crisper picture of local areas. In cubble, \code{switch_key()} can be used to create a new grouping level of spatial locations by specifying a clustering variable. Figure \ref{fig:illu-hier} illustrates the relationship of cubbles at station and cluster level, in both the long and nested forms. By specifying \code{cluster_nested <- station_nested \%>\% switch_key(key = cluster)}, the cubble redefines the cubble key from the \code{id} in \code{station_nested} to \code{cluster} into \code{cluster_nested}. All the spatial variables in \code{cluster} are now nested into a \code{.val} column and cluster level variables can be computed in the same fashion as station level variables in \code{station_nested}.

@@

One task that may interest analysts in spatio-temporal data is to find how similar time series, from nearby sites, are. This problem can be seen as a matching problem [@stuart2010matching; @mcintosh2018using] that pairs up similar time series in nearby locations or a data fusing exercise that merges data collected from different sources [@cocchi2019data]. \code{match_sites()} in \pkg{cubble} provides a simple algorithm for this task. The algorithm first matches the two data sources spatially by computing the pairwise distance on latitude and longitude. After that, pairs that pass the spatial matching are then matched temporally by computing the number of matched peaks within a fixed length moving window. Figure \ref{fig:illu-matching} illustrates this temporal matching in more detail. Given two series \code{A} and \code{a}, three peaks have been identified in each series. An interval, with default length of five, is constructed for each peak in series \code{A}, while the peaks in series \code{a} are tested against whether they fall into any of the intervals. In this illustration, there are two matches for these two series. Several arguments are available in \code{match_sites()} to fine-tune the matching:

@@

The linking structure of two forms in cubble fits naturally with the interactive graphic frameworks discussed in the literature [@buja1988elements; @buja1996interactive;  @sutherland2000orca; @xie2014reactive; @cheng2016enabling]. Diagram \ref{fig:illu-interactive} illustrates how linking works from the map to the time series in cubble. The map and time series plot is associated with the nested or long cubble, respectively, and when a user action is captured on the map, the site will be activated in the nested cubble (left). The nested cubble will communicate to the long cubble to activate all the observations with the same \code{id} (middle). The long cubble will then highlight the activated series in the time series plot (right).

### 2022-04-30 arxiv/cubble-arxiv.tex
Advanced section-morework needed

One task that may interest analysts in spatio-temporal data is to find how similar the time series from nearby sites are. This problem can be seen as a matching problem (\protect\hyperlink{ref-stuart2010matching}{Stuart 2010}; \protect\hyperlink{ref-mcintosh2018using}{McIntosh et al. 2018}) that pairs up similar time series in nearby locations or a data fusing exercise that merges data collected from different sources (\protect\hyperlink{ref-cocchi2019data}{Cocchi 2019}). \code{match_sites()} in \pkg{cubble} provides a simple algorithm for this task. The algorithm first matches the two data sources spatially by computing the pairwise distance on latitude and longitude. Pairs that pass the spatial matching are then matched temporally through computing the number of matched peaks within a fixed length window. Figure \ref{fig:illu-matching} illustrates this temporal matching in more details. Given two series \code{A} and \code{a}, three peaks have been picked in each series. An interval, with default length of five, is constructed for each peak in series \code{A} and the peaks in series \code{a} are tested against whether they fall into any of the intervals. In this illustration, there are two matches for these two series. Several arguments are available in \code{match_sites()} to fine-tune the matching:

@@

The linking structure of two forms in cubble fits naturally with the interactive graphic pipeline discussed in the literature (\protect\hyperlink{ref-buja1988elements}{Buja, Asimov, and Hurley 1988}; \protect\hyperlink{ref-buja1996interactive}{Buja, Cook, and Swayne 1996}; \protect\hyperlink{ref-sutherland2000orca}{Sutherland et al. 2000}; \protect\hyperlink{ref-xie2014reactive}{Xie, Hofmann, and Cheng 2014}; \protect\hyperlink{ref-cheng2016enabling}{Cheng, Cook, and Hofmann 2016}). Diagram \ref{fig:illu-interactive} illustrates how linking works from the map to the time series in cubble. The map and time series plot is associated with the nested or long cubble, respectively, and when a user action is captured on the map, the site will be activated in the nested cubble (left). The nested cubble will communicate to the long cubble to activate all the observations with the same \code{id} (middle). The long cubble will then highlight the activated series in the time series plot (right).

>>>>

One task that may interest analysts in spatio-temporal data is to find how similar time series, from nearby sites, are. This problem can be seen as a matching problem (\protect\hyperlink{ref-stuart2010matching}{Stuart 2010}; \protect\hyperlink{ref-mcintosh2018using}{McIntosh et al. 2018}) that pairs up similar time series in nearby locations or a data fusing exercise that merges data collected from different sources (\protect\hyperlink{ref-cocchi2019data}{Cocchi 2019}). \code{match_sites()} in \pkg{cubble} provides a simple algorithm for this task. The algorithm first matches the two data sources spatially by computing the pairwise distance on latitude and longitude. After that, pairs that pass the spatial matching are then matched temporally by computing the number of matched peaks within a fixed length moving window. Figure \ref{fig:illu-matching} illustrates this temporal matching in more detail. Given two series \code{A} and \code{a}, three peaks have been identified in each series. An interval, with default length of five, is constructed for each peak in series \code{A}, while the peaks in series \code{a} are tested against whether they fall into any of the intervals. In this illustration, there are two matches for these two series. Several arguments are available in \code{match_sites()} to fine-tune the matching:

@@

The linking structure of two forms in cubble fits naturally with the interactive graphic frameworks discussed in the literature (\protect\hyperlink{ref-buja1988elements}{Buja, Asimov, and Hurley 1988}; \protect\hyperlink{ref-buja1996interactive}{Buja, Cook, and Swayne 1996}; \protect\hyperlink{ref-sutherland2000orca}{Sutherland et al. 2000}; \protect\hyperlink{ref-xie2014reactive}{Xie, Hofmann, and Cheng 2014}; \protect\hyperlink{ref-cheng2016enabling}{Cheng, Cook, and Hofmann 2016}). Diagram \ref{fig:illu-interactive} illustrates how linking works from the map to the time series in cubble. The map and time series plot is associated with the nested or long cubble, respectively, and when a user action is captured on the map, the site will be activated in the nested cubble (left). The nested cubble will communicate to the long cubble to activate all the observations with the same \code{id} (middle). The long cubble will then highlight the activated series in the time series plot (right).
